---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
[![Build Status](https://travis-ci.org/sa-lee/plyranges.svg?branch=master)](https://travis-ci.org/sa-lee/plyranges)
[![Coverage Status](https://img.shields.io/codecov/c/github/sa-lee/plyranges/master.svg)](https://codecov.io/github/sa-lee/plyranges?branch=master)

# plyranges

`plryanges` provides a consistent interface for importing and
wrangling genomics data from a variety of sources. The package defines a grammar of genomic data  manipulation based on `dplyr` and the Bioconductor packages `IRanges`, `GenomicRanges`, and `rtracklayer`. It does this by providing a
set of verbs for developing analysis pipelines based on _Ranges_ objects
that represent genomic regions:

* Modify genomic regions with the `set_width()` and `stretch()` functions.
* Modify genomic regions while fixing the start/end/center coordinates with the `anchor_` family of functions.
* Sort genomic ranges with `arrange()`.
* Modify, subset, and aggregate genomic data with the `mutate()`, 
`filter()`, and `summarise()`functions.
* Any of the above operations can be performed on partitions of the
data with `group_by()`.
* Find nearest neighbour genomic regions with the `join_nearest_` family
of functions.
* Find overlaps between ranges with the `join_overlaps_` family of functions.
* Merge all overlapping and adjacent genomic regions with `reduce_ranges()`.
* Merge the end points of all genomic regions with `disjoin_ranges()`.
* Import and write common genomic data formats with the `read_/write_` family
of functions.

For more details on the features of plryanges, read the 
[introduction vignette](https://sa-lee.github.io/plyranges/articles/an-introduction.html).

# Installation

Currently only the development version is available via Bioncoductor
```{r, eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("sa-lee/plyranges")
```

# Quick overview

## About `Ranges`

`Ranges` objects can either represent sets of integers as `IRanges`
(which have start, end and width attributes)
or represent genomic intervals (which have additional attributes, sequence name,
and strand) as `GRanges`. 
In addition, both types of `Ranges` can store information
about their intervals as metadata columns (for example GC content 
over a genomic interval). 

`Ranges` objects follow the tidy data principle: each row
of a `Ranges` object corresponds to an interval, while each column will represent
a variable about that interval, and generally each object will represent a
single unit of observation (like gene annotations). 

We can construct a `IRanges` object from a `data.frame` with a `start` or
`width` using the `as_iranges()` method.

```{r, message=FALSE}
library(plyranges)
df <- data.frame(start = 1:5, width = 5)
as_iranges(df)
# alternatively with end
df <- data.frame(start = 1:5, end = 5:9)
as_iranges(df)
```

We can also construct a `GRanges` object in a similar manner. Note that
a `GRanges` object requires at least a seqnames column to be present in the
data.frame (but not necessarily a strand column). 

```{r}
df <- data.frame(seqnames = c("chr1", "chr2", "chr2", "chr1", "chr2"),
                 start = 1:5,
                 width = 5)
as_granges(df)
# strand can be specified with `+`, `*` (mising) and `-`
df$strand <- c("+", "+", "-", "-", "*")
as_granges(df)
```

# Example: finding GWAS hits that overlap known exons
Let's look at a more a realistic example (taken from HelloRanges vignette).

```{r, include=FALSE}
dir <- system.file(package = "HelloRangesData", "extdata/")
genome <- as_granges(read.delim(file.path(dir, "hg19.genome"), 
                     header = FALSE),
                     seqnames = V1, start = 1L, width = V2)

gwas <- read_bed(file.path(dir, "gwas.bed"), genome_info = genome)
exons <- read_bed(file.path(dir, "exons.bed"), genome_info = genome)
```

Suppose we have two _GRanges_ objects: one containing coordinates of known exons
and another containing SNPs from a GWAS.

The first and last 5 exons are printed below, there are two additional
columns corresponding to the exon name, and a score.

We could check the number of exons per chromosome using `group_by` and
`summarise`.
```{r}
exons 
exons %>%
  group_by(seqnames) %>%
  summarise(n = n())
```

Next we create a column representing the transcript_id with `mutate`:

```{r}
exons <- exons %>%
  mutate(tx_id = sub("_exon.*", "", name))
```

To find all GWAS SNPs that overlap exons, we use `join_overlap_inner`. This
will create a new _GRanges_ with the coordinates of SNPs that overlap exons,
as well as metadata from both objects.

```{r}
olap <- join_overlap_inner(gwas, exons)
olap
```

For each SNP we can count the number of times it overlaps a transcript.

```{r}
olap %>%
  group_by(name.x, tx_id) %>%
  summarise(n = n())
```

We can also generate 2bp splice sites on either side of the exon
using `flank_left` and `flank_right`. We add a column indicating the side 
of flanking for illustrative purposes. The `combine_ranges` function pairs
the left and right pairs.

```{r}
left_ss <- flank_left(exons, 2L) %>%
  mutate(side = "left")
right_ss <- flank_right(exons, 2L) %>%
  mutate(side = "right")
all_ss <- combine_ranges(left_ss, right_ss)
all_ss
```

# Learning more

Read the [introduction vignette](https://sa-lee.github.io/plyranges/articles/an-introduction.html) for an overview of `plyranges` features. 

# About the design of plyranges API

The `plyranges` package aims to provide a 
[fluent](https://en.wikipedia.org/wiki/Fluent_interface) interface for 
performing common genomic data analysis tasks. One goal of this package is to 
decrease the learning curve for Bioconductor and S4 classes 
(especially for new users) by providing a consistent interface to 
common classes _IRanges_ and _GRanges_ 
(future work will extend this to the _SummarizedExperiment_  class)

All of the methods defined in `plyranges` are type-consistent and 
chainable. As a consequence, users familiar with 
`dplyr` should be able to read and understand `plyranges` code without 
difficulty.

